---
title: "Mouse pseudotime"
author: "James Ashmore"
date: "2020-04-21"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Setup

### Packages

Set chunk options:

```{r}
knitr::opts_chunk$set(
  autodep    = TRUE,
  cache      = TRUE,
  cache.path = "cache/mouse-pseudotime.Rmd/",
  dev        = "png",
  error      = FALSE,
  message    = FALSE,
  warning    = FALSE
)
```

Load Bioconductor packages:

```{r}
library(BiocParallel)
library(MouseGastrulationData)
library(TSCAN)
library(scater)
library(scran)
library(scuttle)
library(slingshot)
library(tradeSeq)
```

Load CRAN packages:

```{r}
library(RColorBrewer)
library(pheatmap)
```

### Data

Read experiment object:

```{r}
sce <- readRDS("output/mouse-trajectory.rds")
```

Make feature names unique:

```{r}
rownames(sce) <- uniquifyFeatureNames(rowData(sce)$gene_id, rowData(sce)$gene_name)
```

Make cell names unique:

```{r}
colnames(sce) <- seq_len(ncol(sce))
```

Convert slingshot clusters to a factor:

```{r}
sce$slingClusters <- as.factor(sce$slingClusters)
```

Tidy column data for plotting and visualization:

```{r}
dat <- colData(sce)

dat <- cbind(cell = seq_len(ncol(sce)), dat)

ids <- grep("slingPseudotime", colnames(dat), invert = TRUE)

dat <- reshape2::melt(as.data.frame(dat), id.vars = ids, variable.name = "lineage", value.name = "pseudotime")

dat$lineage <- gsub("slingPseudotime_", "", dat$lineage)
```

### Palettes

Define palette colours:

```{r}
EmbryoCelltypeColours <- MouseGastrulationData::EmbryoCelltypeColours

EmbryoGenotypeColours <- c("wt" = "#76B7B2", "het" = "#B07AA1", "hom" = "#EDC948")

EmbryoSourceColours <- c("atlas" = "#BAB0AC", "embryo" = "#E15759")

EmbryoClusterColours <- setNames(brewer.pal(nlevels(sce$slingClusters), name = "Set3"), levels(sce$slingClusters))
```

Define palette labels:

```{r}
EmbryoGenotypeLabels <- c("wt" = "Flk1 (WT)", "het" = "Flk1 (Het)", "hom" = "Flk1 (Hom)")

EmbryoSourceLabels <- c("atlas" = "Embryo Atlas", "embryo" = "Embryo Body")
```

### Functions

Define function to plot heatmap of gene expression values:

```{r}
plotHeatmap <- function(object, features, columns, order_columns_by) {
  
  obj <- object[features, columns]

  mat <- t(scale(t(assay(obj, "reconstructed"))))

  mat <- mat[, order(obj[[order_columns_by]])]

  col <- colorRampPalette(rev(brewer.pal(5, name = "RdBu")))(100)

  lim <- max(abs(mat))
  
  brk <- seq(-3, 3, length.out = 101)
  
  ann <- list(
    
    col = data.frame(
      Celltype  = obj$celltype,
      Genotype  = obj$genotype,
      Source    = obj$source,
      Cluster   = obj$slingClusters,
      row.names = colnames(obj)
    ),
    
    colors = list(
      Celltype = EmbryoCelltypeColours[unique(obj$celltype)],
      Genotype = EmbryoGenotypeColours[unique(obj$genotype)],
      Source   = EmbryoSourceColours[unique(obj$source)],
      Cluster  = EmbryoClusterColours[unique(obj$slingClusters)]
    )
  )

  plt <- pheatmap(
    mat = mat,
    color = col, 
    breaks = brk,
    scale = "none",
    cluster_cols = FALSE,
    annotation_col = ann$col,
    annotation_colors = ann$colors,
    show_colnames = FALSE
  )
  
  return(plt)

}
```

## TSCAN

Create metadata slot to keep test results:

```{r}
metadata(sce)$testPseudotime <- list(slingPseudotime_1 = NULL)
```

### Lineage 1

Test for differences along pseudotime:

```{r}
de1 <- TSCAN::testPseudotime(sce, pseudotime = sce$slingPseudotime_1, df = 5, block = sce$batch)

de1$ensembl <- rowData(sce)$gene_id

de1$symbol <- rowData(sce)$gene_name

de1 <- subset(de1, FDR < 0.05)

de1 <- de1[order(abs(de1$logFC), decreasing = TRUE), ]

metadata(sce)$testPseudotime$slingPseudotime_1 <- de1
```

Display table of significant genes:

```{r}
DT::datatable(
  data = as.data.frame(de1[, c("ensembl", "symbol", "logFC", "p.value", "FDR")]), 
  colnames = c("Ensembl", "Symbol", "FC", "P", "FDR"),
  rownames = FALSE
)
```

Display heatmap of significant genes:

```{r fig.width = 14, fig.height = 9}
plotHeatmap(
  object = sce,
  features = head(rownames(de1), n = 50),
  columns = complete.cases(sce$slingPseudotime_1),
  order_columns_by = "slingPseudotime_1"
)
```

## TradeSeq

### Linear model

Evaluate the optimal number of knots:

```{r eval = FALSE}
set.seed(1001001001)

aic <- evaluateK(counts = counts(sce), sds = SlingshotDataSet(sce))
```

Identify 2000 most variable genes:

```{r}
dec <- metadata(sce)$combineVar

hvg <- getTopHVGs(dec, n = 2000)

idx <- which(rowData(sce)$gene_id %in% hvg)
```

Fit the NB-GAM model to variable genes:

```{r}
fit <- fitGAM(
  counts = counts(sce),
  sds = SlingshotDataSet(sce),
  U = model.matrix(~ batch, sce@colData),
  genes = idx
)
```

### Association test

Test whether average gene expression is associated with pseudotime:

```{r}
res <- associationTest(fit, lineages = TRUE)

res$ensembl <- rowData(sce)$gene_id[idx]

res$symbol <- rowData(sce)$gene_name[idx]
```

Identify genes associated with lineage 1 pseudotime:

```{r}
de1 <- subset(res, p.adjust(pvalue_1, method = "fdr") < 0.05)

de1 <- de1[order(de1$waldStat_1, decreasing = TRUE), ]

DT::datatable(as.data.frame(de1), rownames = FALSE)
```

Save results table of significant genes for downstream analysis:

```{r}
metadata(sce)$associationTest <- list(slingPseudotime_1 = de1)
```

Plot expression of 50 most significant genes associated with lineage 1 pseudotime:

```{r fig.width = 14, fig.height = 9}
plotHeatmap(sce, features = head(rownames(de1), n = 50), columns = complete.cases(sce$slingPseudotime_1), order_columns_by = "slingPseudotime_1")
```

### Differential start vs end test

Identify differentially expressed genes between the start and end of a lineage:

```{r}
res <- startVsEndTest(fit, lineages = TRUE)

res$ensembl <- rowData(sce)$gene_id[idx]

res$symbol <- rowData(sce)$gene_name[idx]
```

Extract significant genes for lineage 1 pseudotime:

```{r}
de1 <- subset(res, p.adjust(pvalue_lineage1, method = "fdr") < 0.05)

de1 <- de1[order(de1$waldStat_lineage1, decreasing = TRUE), ]

DT::datatable(as.data.frame(de1), rownames = FALSE)
```

Save results table of significant genes for downstream analysis:

```{r}
metadata(sce)$startVsEndTest <- list(slingPseudotime_1 = de1)
```

Plot expression of 50 most significant genes for lineage 1 pseudotime:

```{r fig.width = 14, fig.height = 9}
plotHeatmap(sce, features = head(rownames(de1), n = 50), columns = complete.cases(sce$slingPseudotime_1), order_columns_by = "slingPseudotime_1")
```

## Summary

Save experiment object with tradeSeq results:

```{r}
rownames(sce) <- rowData(sce)$gene_id

colnames(sce) <- NULL

saveRDS(sce, file = "output/mouse-pseudotime.rds")
```

Print session information:

```{r}
sessionInfo()
```
